Extended Abstract

### 2021-05-11 paper/paper.tex
Format constraint resolution

\newcommand{\subst}[2]{#1 ~ \func{\triangleleft} ~ #2}

@@

\title{Extended Abstract}
\subtitle{Co-Contextual Type Inference for the Linear \Picalc{} in Agda}

@@

(Variables are black, $\type{\textrm{types are blue}}$, $\constr{\textrm{constructors are olive green}}$, $\func{\textrm{functions are gray}}$, and I am sorry this is not a wee poem.)

@@

  \inferrule { } {\tSplit{\tomega}{x}{y}}
\end{mathpar}

@@

@@

We use a $\func{[\![\_]\!]}$ function to interpret constraints $\eqconstr{S}{T}$ and $\sumconstr{S}{T}{R}$ into their proof counterparts $\tEq{S}{T}$ and $\tSplit{S}{T}{R}$.

@@

We encode type inference as a function taking processes with $n$ free variables and returning a metavariable context $\gamma$, a typing context with $n$ free variables containing metavariables in $\gamma$, and a list of constraints on those metavariables $\gamma$:
$$
\func{inferProc} : \sProc{n} \to ~ \exists \gamma. ~ \tCtx{n}{\gamma} \times \tConstrs{\gamma}
$$
We define a similar function for type inference on expressions, this time also returning a type $\tau$ of type $\ttype{\gamma}$ for the expression.
$$
\func{inferExpr} : \sExpr{n} \to ~ \exists \gamma. ~ \tCtx{n}{\gamma} \times \ttype{\gamma} \times \tConstrs{\gamma}
$$
Let us provide a couple of examples illustrating how this function is defined:
\begin{itemize}
\item
  For the expression $\sfst~e$ we first recursively infer $e$: $\func{inferExpr}~e = \gamma,\Gamma,\tau,Cs$.
  We then add two fresh metavariables $?t$ and $?s$ of kind $\ktype$ to $\gamma$, add constraints $\eqconstr{\tau}{\tprod~?t~?s}$ and $\sumconstr{?s}{?s}{?s}$ to $cs$, then return $?t$ as the type of $e$.

\item
  For the expression $\scase~e~p~q$ we first recursively infer $p$ and $q$ and $e$: $\func{inferProc}~p = \theta_p, (\Theta_p, s) , Cs_p$, $\func{inferProc}~q = \theta_q, (\Theta_q, t) , Cs_q$, and $\func{inferExpr}~e = \delta_e, \Delta_e, \tau, Cs_e$.
  We create a new metavariable context $\gamma$ and a new typing context $\Gamma$ where every type is a fresh metavariable in $\gamma$.
  We take the union of the metavariables in $\gamma$, $\theta_p$, $\theta_q$ and $\delta_e$
  We take the union of the constraints in $Cs_p$, $Cs_q$ and $Cs_e$, and create new constraints $\eqconstr{\Theta_p}{\Theta_q}$, $\sumconstr{\Gamma}{\Delta_e}{\Theta_p}$, and $\eqconstr{\tau}{\tprod~s~t}$ and return $\Gamma$ as the inferred context.
\end{itemize}

@@

Every substitution \(\sigma\) that makes the constraints hold will make
the process typable under the substituted context. \[
\begin{aligned}
\texttt{infer-sound} &: \forall (p : \texttt{Proc}~n). ~ \texttt{infer}~P \equiv \gamma , \Gamma , cs \\
&\to \forall (\sigma : \texttt{Subst}~\gamma~\delta). ~ [\![ \sigma \triangleleft cs ]\!] \\
&\to (\sigma \triangleleft \Gamma) \vdash P
\end{aligned}
\]

@@

For every context \(\Delta\) that makes the process typable there exists
a most general substitution \(\sigma\) that will solve the constraints
and which substitutes in \(\Gamma\) to something more general than
\(\Delta\). \[
\begin{aligned}
\texttt{infer-complete} &: \forall (p : \texttt{Proc}~n). ~ \texttt{infer}~P \equiv \gamma , \Gamma , cs \\
&\to \forall (\Delta : \texttt{Ctx}~n~\gamma). ~ \Delta \vdash P \\
&\to \exists \delta. ~ \exists (\sigma : \texttt{Subst}~\gamma~\delta). \\
&\to [\![ \sigma \triangleleft cs ]\!] \times \Delta \subseteq (\sigma \triangleleft \Gamma)
\end{aligned}
\] where \[
\Delta \subseteq \Gamma \triangleq \exists \delta. ~ \exists (\sigma : \texttt{Subst} \_ \delta). ~ \Delta \equiv (\sigma \triangleleft \Gamma)
\]

\hypertarget{constraint-resolution}{%
\section{Constraint Resolution}\label{constraint-resolution}}

@@

Given a set of accumulated substitutions and some constraints, we will
return (hopefully) some further substitutions, and a set of leftover
constraints to which the substitutions have been applied. These leftover
constraints can either not be simplified or are unsatisfiable. \[
\begin{aligned}
\texttt{solve} &: \texttt{Subst}~\gamma~\delta \times \texttt{List}~(\texttt{Constr}~\delta) \\
&\to \texttt{Subst}~\gamma~\phi \times \texttt{List}~(\texttt{Constr}~\phi)
\end{aligned}
\]

@@

The simplified constraints are enough to entail the original
constraints. \[
\begin{aligned}
\texttt{solve-sound}
&: \texttt{solve}~(\sigma_1, cs_1) \equiv (\sigma_2, cs_2) \\
&\to \forall (\sigma_f : \texttt{Subst}~\gamma~\delta) \\
& \to [\![ \sigma_f \triangleleft cs_2 ]\!] \to [\![ \sigma_f \cdot \sigma_2 \triangleleft cs_1 ]\!]
\end{aligned}
\]

@@

Any substitution \(\sigma_f\) that makes the original constraints
\(cs_1\) hold will be a specialisation \(\sigma_f\) of our returned
substitution \(\sigma_2\). The specialisation \(\sigma_g\) will make the
returned constraints \(cs_2\) hold. \[
\begin{aligned}
\texttt{solve-sound}
&: \texttt{solve}~(\sigma_1, cs_1) \equiv (\sigma_2, cs_2) \\
&\to \forall (\sigma_f : \texttt{Subst}~\gamma~\delta). ~ [\![ \sigma_f \triangleleft cs_1 ]\!] \\
&\to \exists (\sigma_g : \sigma_f = \sigma_g \cdot \sigma_2) \times [\![ \sigma_g \triangleleft cs_2 ]\!]
\end{aligned}
\]

@@

Nothing prevents us from returning the original constraints as output.
We therefore promise that every constraint we return is either currently
unsolvable because we have not enough information, or is unsatisfiable
altogether. \[
\begin{aligned}
\texttt{solve-sound}
&: \texttt{solve}~(\sigma_1, cs_1) \equiv (\sigma_2, cs_2) \\
&\to \forall c \in cs_2. ~ \texttt{IsDeferred}~c \uplus \texttt{IsUnsat}~c
\end{aligned}
\] where \[
\texttt{IsUnsat}~c = \forall \sigma. ~ \neg [\![ \sigma \triangleleft c ]\!]
\]

@@

\[
\forall c. ~ \texttt{IsDeferred}~c \to \exists \sigma. [\![ \sigma \triangleleft c ]\!]
\]

>>>>

\newcommand{\subst}[2]{#1 ~ \func{\triangleleft} ~ #2}
\newcommand{\tSubst}[2]{\type{Subst}~#1~#2}
\newcommand{\interpr}[1]{\func{[\![} #1 \func{]\!]}}

@@

\title
[Co-Contextual Type Inference for the Linear \Picalc{} in Agda]
{Co-Contextual Type Inference \\ for the Linear \Picalc{} \\ in Agda}
\subtitle{(Extended Abstract)}

@@

(Variables are black, $\type{\textrm{types are blue}}$, $\constr{\textrm{constructors are olive green}}$, $\func{\textrm{functions are gray}}$, and we are sorry this is not a wee poem.)

@@

  \inferrule { } {\tSplit{\tomega}{x}{y}}
\end{mathpar}
\todo{generalisation to an algebra?}

@@

\todo{Introduce Subst type}

@@

We use a $\interpr{\_}$ function to interpret constraints $\eqconstr{S}{T}$ and $\sumconstr{S}{T}{R}$ into their proof counterparts $\tEq{S}{T}$ and $\tSplit{S}{T}{R}$.

@@

We encode type inference as a function taking processes with $n$ free variables and returning a metavariable context $\gamma$, a typing context with $n$ free variables containing metavariables in $\gamma$, and a list of constraints on those metavariables $\gamma$:
$$
\func{inferProc} : \sProc{n} \to ~ \exists \gamma. ~ \tCtx{n}{\gamma} \times \tConstrs{\gamma}
$$
We define a similar function for type inference on expressions, this time also returning a type $\tau$ of type $\ttype{\gamma}$ for the expression.
$$
\func{inferExpr} : \sExpr{n} \to ~ \exists \gamma. ~ \tCtx{n}{\gamma} \times \ttype{\gamma} \times \tConstrs{\gamma}
$$
This functions are total: if processes and expressions are untypable, their constraints will simply be unsolvable.
Let us provide a couple of examples illustrating how this function is defined:
\begin{itemize}
\item
  For the expression $\sfst~e$ we first recursively infer $e$: $\func{inferExpr}~e = \gamma,\Gamma,\tau,Cs$.
  We then add two fresh metavariables $?t$ and $?s$ of kind $\ktype$ to $\gamma$, add constraints $\eqconstr{\tau}{\tprod~?t~?s}$ and $\sumconstr{?s}{?s}{?s}$ to $cs$, then return $?t$ as the type of $e$.

\item
  For the expression $\scase~e~p~q$ we first recursively infer $p$ and $q$ and $e$: $\func{inferProc}~p = \theta_p, (\Theta_p, s) , Cs_p$, $\func{inferProc}~q = \theta_q, (\Theta_q, t) , Cs_q$, and $\func{inferExpr}~e = \delta_e, \Delta_e, \tau, Cs_e$.
  We create a new metavariable context $\gamma$ and a new typing context $\Gamma$ where every type is a fresh metavariable in $\gamma$.
  We take the union of the metavariables in $\gamma$, $\theta_p$, $\theta_q$ and $\delta_e$
  We take the union of the constraints in $Cs_p$, $Cs_q$ and $Cs_e$, and create new constraints $\eqconstr{\Theta_p}{\Theta_q}$, $\sumconstr{\Gamma}{\Delta_e}{\Theta_p}$, and $\eqconstr{\tau}{\tsum~s~t}$ and return $\Gamma$ as the inferred context.
\end{itemize}

@@

Every substitution \(\sigma\) that makes the constraints hold will make the process typable under the substituted context.
\begin{flalign*}
& \tEq{\func{infer}~p}{\gamma , \Gamma , cs} \to (\sigma : \tSubst{\gamma}{\delta}) \to \interpr{\subst{\sigma}{cs}} \to \tProc{(\subst{\sigma}{\Gamma})}{p} &&
\end{flalign*}

@@

For every context \(\Delta\) that makes the process typable there exists a most general substitution \(\sigma\) that will solve the constraints and which substitutes in \(\Gamma\) to something more general than \(\Delta\).
\begin{flalign*}
& \tEq{\func{infer}~p}{\gamma , \Gamma , cs} \to (\Delta : \tCtx{n}{\gamma}) \to \tProc{\Delta}{p} && \\
&\to \exists \delta. ~ \exists (\sigma : \texttt{Subst}~\gamma~\delta). \interpr{\subst{\sigma}{cs}} \times (\tProc{\subst{\sigma}{\Gamma}}{p}) \times \Delta \subseteq (\subst{\sigma}{\Gamma}) &&
\end{flalign*}
where $$
\Delta \func{\subseteq} \Gamma \triangleq \exists \delta. ~ \exists (\sigma : \tSubst{\_}{\delta}). ~ \tEq{\Delta}{(\subst{\sigma}{\Gamma})}
$$

\section{Constraint Resolution}\label{constraint-resolution}

@@

Given a set of accumulated substitutions and some constraints, we will
return (hopefully) some further substitutions, and a set of leftover
constraints to which the substitutions have been applied. These leftover
constraints can either not be simplified or are unsatisfiable. \[
\begin{aligned}
\func{solve} &: \tConstrs{\gamma} \to \tSubst{\gamma}{\delta} \times \tConstrs{\delta}
\end{aligned}
\]

@@

The simplified constraints are enough to entail the original constraints.
\[
\begin{aligned}
& ~ \tEq{\func{solve}~cs_1}{(\sigma, cs_2)} \to (\sigma_f : \tSubst{\gamma}{\delta}) \\
\to& \interpr{\subst{\sigma_f}{cs_2}} \to \interpr{\subst{\sigma_f \cdot \sigma}{cs_1}}
\end{aligned}
\]

@@

Any substitution \(\sigma_f\) that makes the original constraints \(cs_1\) hold will be a specialisation \(\sigma_f\) of our returned substitution \(\sigma_2\). The specialisation \(\sigma_g\) will make the returned constraints \(cs_2\) hold.
\begin{flalign*}
&\tEq{\func{solve}~cs_1}{(\sigma, cs_2)} \to (\sigma_f : \tSubst{\gamma}{\delta}) \to \interpr{\subst{\sigma_f}{cs_1}} && \\
&\to \exists (\sigma_g : \tSubst{\_}{\delta}). ~ \tEq{\sigma_f}{\sigma_g \cdot \sigma} \times \interpr{\subst{\sigma_g}{cs_2}} &&
\end{flalign*}

@@

Nothing prevents us from returning the original constraints as output.
We therefore promise that every constraint we return is either currently
unsolvable because we have not enough information, or is unsatisfiable
altogether.
\begin{flalign*}
&\func{solve}~cs_1 \equiv (\sigma, cs_2) \to \forall c \in cs_2. ~ \type{IsDeferred}~c ~ \type{\uplus} ~ \type{IsUnsat}~c &&
\end{flalign*}
where
\begin{flalign*}
& \type{IsUnsat}~c \triangleq \forall \sigma. ~ \neg \interpr{\subst{\sigma}{c}} &&
\end{flalign*}

@@

\begin{flalign*}
& \forall c. ~ \type{IsDeferred}~c \to \exists \sigma. \interpr{\subst{\sigma}{c}} &&
\end{flalign*}
